---
title: "Comparative Transcriptomic"
output: pdf_document
date: "2026-01-13"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
```

## Comparative Transcriptomic - Naive vs Memory fasting response
```{r}
memory.fasted_vs_fed <- as.data.frame(readRDS("MetID28_res_memory.RDS"))
naive.fasted_vs_fed <- as.data.frame(readRDS("MetID28_res_naive.RDS"))
memory.fasted_vs_fed$geneID <- rownames(memory.fasted_vs_fed)
naive.fasted_vs_fed$geneID <- rownames(naive.fasted_vs_fed)

head(memory.fasted_vs_fed)
head(naive.fasted_vs_fed)
dim(memory.fasted_vs_fed)
```
```{r}
padj_T <- 0.01

memory_all_directedpadj <- -log10(memory.fasted_vs_fed$padj)*sign(memory.fasted_vs_fed$log2FoldChange)
naive_all_directedpadj <- -log10(naive.fasted_vs_fed$padj)*sign(naive.fasted_vs_fed$log2FoldChange)

res_comp <- data.frame(memory_dirpVal = memory_all_directedpadj, 
                       naive_dirpVal = naive_all_directedpadj, 
                       memory_l2fc = memory.fasted_vs_fed$log2FoldChange,
                       memory_padj = memory.fasted_vs_fed$padj,
                       naive_l2fc = naive.fasted_vs_fed$log2FoldChange,
                       naive_padj = naive.fasted_vs_fed$padj)

res_comp$padj_comp <- "NS"
res_comp$padj_comp[res_comp$memory_padj < padj_T & res_comp$naive_padj < padj_T]<- "sign_memo_and_naive"
res_comp$padj_comp[res_comp$memory_padj < padj_T  & res_comp$naive_padj > padj_T]<- "sign_memo"
res_comp$padj_comp[res_comp$naive_padj < padj_T & res_comp$memory_padj > padj_T]<- "sign_naive"

rownames(res_comp) <- memory.fasted_vs_fed$geneID

res_comp <- na.omit(res_comp)
head(res_comp)
```

## Directed pValue : Memory against Naive
```{r}
ggplot(res_comp, aes(x = memory_dirpVal, y = naive_dirpVal)) +
  geom_abline(aes(intercept = 0, slope = 1), size = 0.7,linetype = 'dashed') +
  geom_point(aes(fill = padj_comp), color = "black", shape = 21, size = 2, alpha = 0.5, stroke = 0.3) +
  scale_fill_manual(name = "",values = c("#d9d9d9", "#8c96c6", "#810f7c", "dodgerblue4", "#d9d9d9"),
                     labels = c("not significant", 
                                "significant in x direction",
                                "significant in both directions",
                                "significant in y direction"))+
  coord_fixed(ratio = 1, xlim = NULL, ylim = NULL, expand = TRUE, clip = "on")+
  theme_minimal() +
  labs(x = "Directed pValue \nMemory fasted vs fed",
       y = "Directed pValue \nNaive fasted vs fed",
       title = "Comparative transcriptomic")
ggsave("images/MetID28.comparative_trans.directed_pvalue.png")
```
## L2FC pValue : Memory against Naive
  geom_abline(aes(intercept = 0, slope = 1, colour = "x = y"), size = 0.7) +
  geom_smooth(aes(colour = "Linear regression"), method = "lm", se = FALSE, size = 0.7) +
    scale_colour_manual(name = "Lines", values = c("x = y" = "black", "Linear regression" = "blue")) +

```{r}

ggplot(res_comp, aes(x = memory_l2fc, y = naive_l2fc)) +
  geom_abline(aes(intercept = 0, slope = 1, colour = "x = y"), size = 0.7,linetype = 'dashed') +
  geom_vline(xintercept = c(-0.3, 0.3), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = c(-0.3,0.3), col = "gray", linetype = 'dashed')+
  geom_point(aes(fill = padj_comp), color = "black", shape = 21, size = 2, alpha = 0.7, stroke = 0.3) +
  scale_fill_manual(name = "",values = c("#d9d9d9", "#8c96c6", "#810f7c", "dodgerblue4", "#d9d9d9"),
                     labels = c("not significant", 
                                "significant in x direction",
                                "significant in both directions",
                                "significant in y direction"))+
                coord_fixed( ratio=1) +
theme_minimal() +
  labs(x = "l2fc \nMemory fasted vs fed",
       y = "l2fc  \nNaive fasted vs fed",
       title = "Comparative l2fc")
ggsave("images/MetID28.comparative_trans.L2FC.png")

```
```{r}
# calcul des limites communes
lims <- range(c(res_comp_sig$memory_l2fc, res_comp$naive_l2fc), na.rm = TRUE)

ggplot(res_comp_sig, aes(x = memory_l2fc, y = naive_l2fc)) +
  geom_smooth(aes(colour = "Linear regression"), method = "lm", se = FALSE, size = 0.7) +
  geom_abline(aes(intercept = 0, slope = 1, colour = "x = y"),size = 0.7, linetype = 'dashed') +
  geom_vline(xintercept = c(-0.3, 0.3), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = c(-0.3, 0.3), col = "gray", linetype = 'dashed') +
  geom_point(aes(fill = padj_comp),
             color = "black", shape = 21, size = 2, alpha = 0.7, stroke = 0.3) +
  scale_fill_manual(
    name = "",
    values = c("#8c96c6", "#810f7c", "dodgerblue4", "#d9d9d9","black","red"),
    labels = c("significant in x direction",
               "significant in both directions",
               "significant in y direction")
  ) +
  
  scale_x_continuous(limits = lims) +
  scale_y_continuous(limits = lims) +
  
  coord_fixed() +
  theme_minimal() +
  labs(
    x = "l2fc \nMemory fasted vs fed",
    y = "l2fc  \nNaive fasted vs fed",
    title = "Comparative l2fc"
  )
ggsave("images/MetID28.comparative_trans.sig.L2FC-lm.png")
```

```{r}
res_comp_sig <- res_comp[!res_comp$padj_comp == "NS",] 

ggplot(res_comp_sig, aes(x = memory_l2fc, y = naive_l2fc)) +
  
  geom_abline(aes(intercept = 0, slope = 1, colour = "x = y"), size = 0.7,linetype = 'dotted') +
  geom_vline(xintercept = c(-0.3, 0.3), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = c(-0.3,0.3), col = "gray", linetype = 'dashed')+
  geom_point(aes(fill = padj_comp), color = "black", shape = 21, size = 2, alpha = 0.7, stroke = 0.3) +
  scale_fill_manual(name = "",values = c("#8c96c6", "#810f7c", "dodgerblue4", "#d9d9d9"),
                     labels = c("significant in x direction",
                                "significant in both directions",
                                "significant in y direction"))+
  theme_minimal() +
  labs(x = "l2fc \nMemory fasted vs fed",
       y = "l2fc  \nNaive fasted vs fed",
       title = "Comparative l2fc - significant padj")
ggsave("images/MetID28.comparative_trans.L2FCsig.png")
```
### Testing if there is a shift in memory response 
Null hypothesis : The observations 
X−Y are symmetric about μ = 0

## Testing Wilcoxon on sig DEgs in at least one condition
```{r}
#selecting genes that are significant in at least one condition (naive|memory) - and L2FC strong in at least one cdt
res_comp_sig <- res_comp_sig[abs(res_comp_sig$memory_l2fc)>0.3 | abs(res_comp_sig$naive_l2fc)>0.3,]

wilcox.test(res_comp_sig$memory_l2fc, res_comp_sig$naive_l2fc, paired = TRUE)
```

## Testing Wilcoxon on sig DEgs in both conditions
Test de Wilcoxon apparié : Il nécessite l'hypothèse que la différence entre les échantillons appariés suit une distribution symétrique autour d'un centre. Le but étant de tester si ce centre est 0.
```{r}
#selecting genes that are significant in both condition (naive|memory)
res_comp_sigboth <- res_comp_sig[res_comp_sig$padj_comp=="sign_memo_and_naive",]
res_comp_sigboth <- res_comp_sigboth[abs(res_comp_sigboth$memory_l2fc)>0.3 | abs(res_comp_sigboth$naive_l2fc)>0.3,]
wilcox.test(res_comp_sigboth$memory_l2fc, res_comp_sigboth$naive_l2fc, paired = TRUE)
```
```{r}
wilcox.test(abs(res_comp_sigboth$memory_l2fc-res_comp_sigboth$naive_l2fc), alternative = "greater")
```
Alternative hypothesis - One sided H : The observations 
The observations 
abs(l2fc_memory−l2fc_naive) are symmetric about μ > 0

```{r}
ggplot(res_comp, aes(x = memory_l2fc, y = naive_l2fc)) +
  geom_abline(aes(intercept = 0, slope = 1, colour = "x = y"), size = 0.7, linetype = 'dashed') +
  geom_vline(xintercept = c(-0.3, 0.3), col = "gray", linetype = 'dashed') +
  geom_hline(yintercept = c(-0.3, 0.3), col = "gray", linetype = 'dashed') +
  geom_point(aes(fill = padj_comp), color = "black", shape = 21, size = 2, alpha = 0.7, stroke = 0.3) +
  scale_fill_manual(name = "", values = c("#d9d9d9", "#8c96c6", "#810f7c", "dodgerblue4", "#d9d9d9"),
    labels = c("not significant", 
               "significant in x direction",
               "significant in both directions",
               "significant in y direction")) +
  
  theme_minimal() +
  labs( x = "l2fc \nMemory fasted vs fed", y = "l2fc  \nNaive fasted vs fed", title = "Comparative l2fc")

```


### Apply threshold
```{r}
naive.fasted_vs_fed <- na.omit(naive.fasted_vs_fed)
memory.fasted_vs_fed <- na.omit(memory.fasted_vs_fed)
```

```{r}
padj <- 0.01 
L2FC <- 0.3 # FC of 1.2

naive.fasted_vs_fed <- naive.fasted_vs_fed[naive.fasted_vs_fed$padj < padj 
                                           & abs(naive.fasted_vs_fed$log2FoldChange) > L2FC,]
memory.fasted_vs_fed <- memory.fasted_vs_fed[memory.fasted_vs_fed$padj < padj 
                                           & abs(memory.fasted_vs_fed$log2FoldChange) > L2FC,]
```

### Filter genes that are present in naive and memory response 
Compared to their own reference : memo is 2FC whie naive is 1FC
```{r}
memory.fasted_vs_fed$geneID <- rownames(memory.fasted_vs_fed)
naive.fasted_vs_fed$geneID <- rownames(naive.fasted_vs_fed)

commonGenes <- memory.fasted_vs_fed$geneID[memory.fasted_vs_fed$geneID %in% naive.fasted_vs_fed$geneID]
length(commonGenes) #2046
```

### Compute directed pvalue
-log10(padj)*L2FC
```{r}
memory.fasted_vs_fed$padj.minuslog10 <- -log10(memory.fasted_vs_fed$pvalue)
memory.fasted_vs_fed$directed.padj <- memory.fasted_vs_fed$padj.minuslog10*memory.fasted_vs_fed$log2FoldChange

memory_directed.padj <- memory.fasted_vs_fed$directed.padj[memory.fasted_vs_fed$geneID %in% commonGenes]

naive.fasted_vs_fed$padj.minuslog10 <- -log10(naive.fasted_vs_fed$pvalue)
naive.fasted_vs_fed$directed.padj <- naive.fasted_vs_fed$padj.minuslog10*naive.fasted_vs_fed$log2FoldChange

naive_directed.padj <- naive.fasted_vs_fed$directed.padj[naive.fasted_vs_fed$geneID %in% commonGenes]
```

```{r}
naive_cl2FC <- naive.fasted_vs_fed$log2FoldChange[naive.fasted_vs_fed$geneID %in% commonGenes]
memory_cl2FC <- memory.fasted_vs_fed$log2FoldChange[memory.fasted_vs_fed$geneID %in% commonGenes]

```

```{r}
data <- data.frame( naive_dirpadj = naive_directed.padj,memory_dirpadj = memory_directed.padj, memory_L2FC = memory_cl2FC, naive_L2FC = naive_cl2FC)
rownames(data) <- commonGenes
```

```{r}
ggplot(data, aes(x = memory, y = naive))+
  geom_abline(intercept = 0, slope = 1)+
  geom_point(size = 1.5,alpha=0.3) +
  theme_minimal()+
  ggtitle("Directed pvalue : log10(padj).L2FC")
ggsave("images/MetID28.comparative_trans.DirectedPval_log10padj_x_L2FC.png")
```
```{r}
ggplot(data, aes(x = memory_L2FC, y = naive_L2FC))+
  geom_abline(intercept = 0, slope = 1)+
  geom_smooth(method='lm')+
  geom_point(size = 1.5,alpha=0.3) +
  theme_minimal()+
  ggtitle("Sig common DEgenes in fasting response - L2FC")
ggsave("images/MetID28.comparative_trans.L2FC.png")
```



